---
title: "Density dependence"
---

```{r}
#| label: setup
#| include: false
source("R/utils.R")
source("R/figures.R")
```

## Overview {#sec-topic2-overview}

## Limits to growth: the logistic model {#sec-logistic}

## Density dependence in the real world {#sec-density-dependence}

## The logistic map

The **logistic map** is the simplest discrete-time model that produces both stable dynamics and chaos:

$$
N_{t+1} = r \, N_t (1 - N_t)
$$

where $N_t \in [0, 1]$ is the population scaled by its maximum and $r \in [0, 4]$ is the growth parameter. For $r < 1$ the population collapses to zero; for $1 < r < 3$ it converges to a stable equilibrium $N^* = 1 - 1/r$; beyond $r = 3$ the equilibrium destabilises through a cascade of period-doubling bifurcations that culminates in deterministic chaos near $r \approx 3.57$.

```{r}
#| label: logistic-grid
#| include: false
n_r    <- 400L
n_burn <- 500L
n_keep <- 100L

r_vals <- seq(1, 4, length.out = n_r)

bif_grid <- purrr::map_dfr(seq_len(n_r), function(idx) {
  r <- r_vals[idx]; N <- 0.5
  for (j in seq_len(n_burn)) N <- r * N * (1 - N)
  Ns <- numeric(n_keep)
  for (j in seq_len(n_keep)) { N <- r * N * (1 - N); Ns[j] <- round(N, 5) }
  tibble(r = round(r, 4), N = Ns)
})

ojs_define(bif_grid = bif_grid)
```

::: {.content-visible when-format="html"}
```{ojs}
viewof r_val  = Inputs.range([1, 4],       {step: 0.01, value: 2,   label: "r"})
viewof n0_val = Inputs.range([0.01, 0.99], {step: 0.01, value: 0.5, label: "N₀"})

bif_arr = transpose(bif_grid)

n_ts = 100
ts_data = {
  const pts = [{t: 0, N: n0_val}]
  let N = n0_val
  for (let t = 1; t <= n_ts; t++) {
    N = r_val * N * (1 - N)
    pts.push({t, N})
  }
  return pts
}

ts_plot = Plot.plot({
  width: 290, height: 250,
  x: {label: "t"},
  y: {domain: [0, 1], label: "Nₜ"},
  title: `N₀ = ${n0_val.toFixed(2)},  r = ${r_val.toFixed(2)}`,
  marks: [
    Plot.line(ts_data, {x: "t", y: "N", stroke: "#2980b9", strokeWidth: 1.2}),
    Plot.dot(ts_data,  {x: "t", y: "N", r: 1.2, fill: "#2980b9"})
  ]
})

bif_plot = Plot.plot({
  width: 410, height: 250,
  x: {domain: [1, 4], label: "r", ticks: [1, 2, 3, 4]},
  y: {domain: [0, 1], label: "Nₜ (attractor)"},
  title: "Bifurcation diagram",
  marks: [
    Plot.dot(bif_arr,   {x: "r", y: "N", r: 0.5, fill: "#666", opacity: 0.25}),
    Plot.ruleX([r_val], {stroke: "#e74c3c", strokeWidth: 2})
  ]
})

html`<div style="display:flex; gap:16px; align-items:flex-start; flex-wrap:wrap">
  <div>${ts_plot}</div>
  <div>${bif_plot}</div>
</div>`
```
:::

::: {.content-visible when-format="pdf"}
```{r}
#| label: fig-logistic-bif-static
#| fig-cap: "Bifurcation diagram of the logistic map. For each value of $r$, the attractor values of $N_t$ are plotted after discarding a 500-step transient. Interactive version: <<PUBLISHED URL>>#sec-logistic."
#| echo: false
plot_logistic_map_bif()
```

```{r}
#| label: fig-logistic-ts-static
#| fig-cap: "Logistic map dynamics $N_t$ starting from $N_0 = 0.5$ at four representative values of $r$, illustrating fixed-point, two-cycle, and chaotic regimes."
#| echo: false
plot_logistic_ts_facet()
```
:::


